Jane, John … Leslie? A Historical Method for Algorithmic Gender Prediction

### 2015-08-21 article.Rmd
Revisions to AHR section

In a 2005 report for the American Historical Association, Elizabeth Lunbeck acknowledged "a sea change in the [historical] profession with respect to gender" before going on to describe the limits to this change for female historians, who face ongoing personal discrimination, lower salaries, and barriers to securing high-ranking positions. Lunbeck's report drew in part on a survey of 362 female historians that produced a rich source of responses detailing the deep and multi-faceted challenges facing women in the profession [Lunbeck 2005]. What follows is a quantitative supplement to Lunbeck's analysis that uses our program to analyze gender representation amongst historians across a much larger scale and a much longer time period. We study one of the bedrocks of the historical profession --- scholarly research --- and focus on two types of output: the dissertation and the monograph. Although women have achieved rough parity with men in terms of the number of completed dissertations, female historians continue to face an unequal playing field in the monographic landscape. Applying our method to the pages of the *American Historical Review* reveals a persistent and significant gap in the number of male-authored and female-authored books reviewed by one of the field’s premier academic journals.

We begin with the history dissertation, often the defining scholarly output of a historian's early career. The completion of a dissertation marks a standardized moment of transition out of the "training" phase of the historical profession. To identify how many women and men completed PhD-level training in history each year, we used data supplied by ProQuest for roughly eighty thousand PhD dissertations completed in the United States between 1950 and 2012.^[ProQuest's dataset does not permit us to limit dissertations just to dissertations produced in history departments. These dissertations include dissertations *about* history. For a justification of this approach, see a [series of blog posts](http://lincolnmullen.com/research/history-dissertations/) exploring the dataset. The American Historical Association's dataset of dissertations relies on reporting from departments and does not have the chronological range of the ProQuest data. Nevertheless, applying the same method to the AHA's dataset does not provide any reason for thinking that the proportion of authorship by gender over time is substantially different when limited to just history department dissertations. American Historical Association, *Directory of History Dissertations*, https://secure.historians.org/pubs/dissertations/. Data export provided to author by Liz Townsend, March 3, 2014.] Our program used the year a historian wrote his or her dissertation to estimate a period for when they might have been born, assuming that a historian was between 25 and 45 years old when completing their dissertation. With this temporal information, we were able to infer their gender and chart how the larger representation of women and men changed year-by-year.

@@

A few caveats are in order. First, a book that is never reviewed by the *AHR* may still go on to have a significant impact on the profession. Second, a book's appearance in the *AHR* is not necessarily correlated with the *quality* of the book, as the journal prints negative reviews as well as positive ones. Nevertheless, when a book appears in the *AHR* it serves as a signal that other historians in positions of power have "approved" it for consideration. Even if it garners a negative review, the fact that it appears in the journal at all is a measure of the fact that the profession's gatekeepers have deemed it important enough to review.  It is precisely this professional gatekeeping dimension that we can analyze using the Gender package: are women more likely to be barred from this gate than men, and if so, by how much?

@@

Not every historian wants or seeks a career path that neatly moves from writing a dissertation to authoring a monograph to being reviewed in the *American Historical Review*. For those that do, however, each of these subsequent stages represents a winnowing process: not all graduate students produce dissertations, not all PhD’s write books, not all books are submitted to the *American Historical Review*, and not all submissions are accepted by the *AHR* editors for review.^[As of 2015 the journal's website reminds readers that "the *AHR* receives over 3,000 books a year; we have the resources to publish at most 1,000 reviews a year (approximately 200 per issue)." [American Historical Review n.d.]] Slighty fewer women than men go through the initial stage of writing a history dissertation. But by the final round of gatekeeping – a monograph reviewed in the *AHR* – the gender gap grows until there are little more than half as many women as men. The *American Historical Review* is, unfortunately, far from alone. Many of the profession's most prestigious annual honors go to drastically more men than women. As of 2015, the Pulitzer Prize for History has been awarded to exactly two female historians in the past twenty years ["The Pulitzer Prizes: History" 2015-]. The Bancroft Prize in American History has been awarded to almost five times as many male historians as female historians over that same period ["Bancroft Prize"].

Examples like the Pulitzer and Bancroft Prizes are only the most glaring cases of gender inequity in the historical profession. Overt discrimination still exists, but inequity operates in ways that are far more cloaked and far more complex, making them difficult to recognize and remedy. They take the form of subconscious bias that subtly weakens the assessment of work done by female historians, whether in a seminar paper or a job application. Moreover, gender discrimination frequently intersects with other kinds of racial and socioeconomic inequity that erects even higher barriers in the women's careers [Townsend 2008]. Gendered family dynamics add another layer of challenges for women: a 2011 survey by the American Historical Association revealed that female historians dedicated substantially more time to child and elder care than their male colleagues, leaving them with less time for the kind of academic research that the profession rewards most highly [Townsend 2012]. All of these factors combine and amplify each other to contribute to the gender gap in the historical profession on display in the pages of the *American Historical Review*. The hidden, murky, and systemic dimensions of gender inequity makes methods like the Gender package all the more imperative for reasearchers.

Finally, we used the Gender package to reveal some of the subtler ways in which gender iniquity operates in the historical profession. After inferring the gender of book authors reviewed by the *AHR*, we were also able to infer the gender of the reviewers themselves. This led us to two findings. First, the story is much the same for reviewers as it is for authors: more than twice as many men as women appear as reviewers in the journal. This is important because reviewers for the *AHR* garner a certain degree of prestige from their work - perhaps not to the same prestige as winning a Pulitzer Prize, but prestigious nonetheless. To become a reviewer, one must possess a Ph.D. or equivalent degree and have published at least one book-length manuscript [American Historical Review n.d.]. *AHR* reviewers are understood to have sufficient experience and command over a given sub-field to evaluate the merits and weaknesses of a new contribution to that literature. Being a reviewer for the journal is, in many ways, a proxy for expertise. Which is a problem. When there are twice as many male "experts" as female "experts" appearing in the *AHR* it imbues our understanding of what constitutes historical expertise.  Or perhaps more accurately, who is and who isn't an "expert" in the field.

The second finding we arrived at when examining the gender of *AHR* reviewers was a better understanding of who reviews which book authors. Once again, the methodology reveals patterns that we might suspect but nonetheless tends to stay hidden from direct sight. About three times as many men as women write reviews of male-authored books in the *AHR*. In the case of female-authored books, the ratio of female-to-male reviewers is closer to 50/50. In short, women are far more likely to write reviews of other women, while men tend to gravitate more towards male authors – who are, of course, already over-represented in the *AHR*.

@@

The findings related to *AHR* reviewers help illustrate the ways in which gender inequity is so insidious. It is not always an overt case of deciding to award a prestigious prize to a male historian; it can be as seemingly innocuous as deciding to write a book review of a male author's book. Countless individual actions like these constitute the core relationships that knit together the profession, and too often these tiny actions benefit men more than women: chairing a panel of mostly male historians, taking on a new male advisee, or assigning mostly male-authored books on a syllabus. On their own they might appear benign - one might argue that male historians simply happen to be interested in similar research topics - but in aggregate they contribute to a broader professional landscape that puts female historians at a disadvantage.

One 2013 analysis of 2,500 recent history Ph.D.'s found that "gender played little role in employment patterns across particular professions and industries" [Wood and Townsend 2013]. Our own analysis reveals a different story, at least in terms of getting a book reviewed in the field's most prestigious journal. The *American Historical Review* continues to publish twice as many reviews of male-authored books as female-authored books, and those same reviews are more than twice as likely to be written by a man than a woman. This disparity is put into even sharper relief when set against the much smaller gender gap between female and male historians who complete dissertations. Women might be getting hired at comparable rates to men, but discriminatory gatekeeping remains alive and well in the profession. The Gender package is one more tool to uncover those practices.

@@

The digital turn has made large datasets widely available, yet in the midst of abundance many humanists continue to deal with the problem of incompleteness and scarcity within their sources. Our method uses *abundant* existing datasets to infer additional information from *scarce* datasets. Its major contribution is to incorporate a historical method, one that takes into account how gendered naming practices have changed over time. This temporal approach provides much improved results over simpler, anachronistic, or ahistorical lookup methods. The Gender package is, of course, far from perfect. By relying on historical datasets collected by state agencies, it reinforces a gender binary and larger power structure that obscures how people may identify themselves and how they experience the world. To borrow a phrase from Miriam Posner, the Gender package does not necessarily meet the "radical, unrealized potential of digital humanities" [Posner 2015]. But it might.

The Gender package is entirely open-source, and we have provided detailed documentation for its code and underlying data so that researchers who wish to build upon it can modify the package accordingly. To take one imagined scenario: Facebook allows its users to select from a spectrum of gender identities or to write in their own prefered identity. Because these identities are associated with the user's first name and date of birth, this is exactly the kind of historical dataset that could be incorporated into the Gender package. Doing so would replace the package's simplistic categories of "Male" and "Female" with more detailed categories like "Female to Male", "Bigender", or "Cisgender Female". Although this imagined scenario would create its own set of challenges, it would represent a major improvement in how the Gender package dealt with the complexity and fluidity of gender identity.^[This imagined scenario runs into its own set of problems. Facebook's business model makes it unlikely to grant researchers open access to its proprietary and closely-guarded datasets. And even if it did grant access to this data, Facebook collected it for particular purposes and in particular ways. Just as government agencies like the Social Security Administration collect data to serve the purposes of the American state, Facebook's data collection more accurately reflects the priorities of a global corporation rather than the lived experience of its users.] By building as much flexibility and transparency into the method as possible, we hope that others can continue to make just these kinds of improvements.

>>>>

In a 2005 report for the American Historical Association, Elizabeth Lunbeck acknowledged "a sea change in the [historical] profession with respect to gender" before going on to describe the limits to this change for female historians, who face ongoing personal discrimination, lower salaries, and barriers to securing high-ranking positions. Lunbeck's report drew in part on a survey of 362 female historians that produced a rich source of responses detailing the deep and multi-faceted challenges facing women in the profession [Lunbeck 2005]. What follows is a quantitative supplement to Lunbeck's analysis that uses our program to analyze gender representation amongst historians across a much larger scale and a much longer time period. We study one of the bedrocks of the historical profession---scholarly research---and focus on two types of output: the dissertation and the monograph. Although women have achieved rough parity with men in terms of the number of completed dissertations, female historians continue to face an unequal playing field in the monographic landscape. Applying our method to the pages of the *American Historical Review* reveals a persistent and significant gap in the number of male-authored and female-authored books reviewed by one of the field’s premier academic journals.

We begin with the history dissertation, often the defining scholarly output of a historian's early career. The completion of a dissertation marks a standardized moment of transition out of the training phase of the historical profession. To identify how many women and men completed PhD-level training in history each year, we used data supplied by ProQuest for roughly eighty thousand PhD dissertations completed in the United States between 1950 and 2012.^[ProQuest's dataset does not permit us to limit dissertations just to dissertations produced in history departments. These dissertations include dissertations *about* history. For a justification of this approach, see a [series of blog posts](http://lincolnmullen.com/research/history-dissertations/) exploring the dataset. The American Historical Association's dataset of dissertations relies on reporting from departments and does not have the chronological range of the ProQuest data. Nevertheless, applying the same method to the AHA's dataset does not provide any reason for thinking that the proportion of authorship by gender over time is substantially different when limited to just history department dissertations. American Historical Association, *Directory of History Dissertations*, https://secure.historians.org/pubs/dissertations/. Data export provided to author by Liz Townsend, March 3, 2014.] Our program used the year a historian wrote his or her dissertation to estimate a period for when they might have been born, assuming that a historian was between 25 and 45 years old when completing their dissertation. With this temporal information, we were able to infer their gender and chart how the larger representation of women and men changed year-by-year.

@@

A few caveats are in order. First, a book that is never reviewed by the *AHR* may still go on to have a significant impact on the profession. Second, a book's appearance in the *AHR* is not necessarily correlated with the *quality* of the book, as the journal prints negative reviews as well as positive ones. Nevertheless, when a book appears in the *AHR* it serves as a signal that other historians in positions of power are taking it seriously. Even if it garners a negative review, the fact that it appears in the journal at all is a measure of the fact that the profession's gatekeepers have deemed it important enough to review. It is precisely this professional gatekeeping dimension that we can analyze using the gender package: are books by men more likely to be reviewed then books by women, and if so, how much more likely?

@@

Not every historian wants or seeks a career path that neatly moves from writing a dissertation to authoring a monograph to being reviewed in the *American Historical Review*. For those that do, however, each of these subsequent stages represents a winnowing process: not all graduate students produce dissertations, not all PhDs write books, not all books are submitted to the *American Historical Review*, and not all submissions are accepted by the *AHR* editors for review.^[As of 2015 the journal's website reminds readers that "the *AHR* receives over 3,000 books a year; we have the resources to publish at most 1,000 reviews a year (approximately 200 per issue)." [American Historical Review n.d.]] Slighty fewer women than men go through the initial stage of writing a history dissertation. But by the final round of gatekeeping---a monograph reviewed in the *AHR*---the gender gap grows until there are little more than half as many women as men. The *American Historical Review* is, unfortunately, far from alone. Many of the profession's most prestigious annual honors go to drastically more men than women. As of 2015, the Pulitzer Prize for History has been awarded to exactly two female historians in the past twenty years ["The Pulitzer Prizes: History" 2015-]. The Bancroft Prize in American History has been awarded to almost five times as many male historians as female historians over that same period ["Bancroft Prize"].

Examples like the Pulitzer and Bancroft Prizes are only the most obvious cases of gender inequity in the historical profession. Overt discrimination still exists, but inequity operates in ways that are far more cloaked and far more complex, making them difficult to recognize and remedy. They take the form of subconscious bias that subtly weakens the assessment of work done by female historians, whether in a seminar paper or a job application. Moreover, gender discrimination frequently intersects with other kinds of racial and socioeconomic inequity that erects even higher barriers in the women's careers [Townsend 2008]. Gendered family dynamics add another layer of challenges for women: a 2011 survey by the American Historical Association revealed that female historians dedicated substantially more time to child and elder care than their male colleagues, leaving them with less time for the kind of academic research that the profession rewards most highly [Townsend 2012]. All of these factors combine and amplify each other to contribute to the gender gap in the historical profession on display in the pages of the *American Historical Review*. The hidden, murky, and systemic dimensions of gender inequity makes methods like the gender package all the more imperative for reasearchers.

Finally, we used the gender package to reveal some of the subtler ways in which gender iniquity operates in the historical profession. After inferring the gender of book authors reviewed by the *AHR*, we were also able to infer the gender of the reviewers themselves. This led us to two findings. First, the story is much the same for reviewers as it is for authors: more than twice as many men as women appear as reviewers in the journal. This is important because reviewers for the *AHR* garner a certain degree of prestige from their work---perhaps not to the same prestige as winning a Pulitzer Prize, but prestige nonetheless. To become a reviewer, one must possess a PhD or equivalent degree and have published at least one book-length manuscript [American Historical Review n.d.]. *AHR* reviewers are understood to have sufficient experience and command over a given sub-field to evaluate the merits and weaknesses of a new contribution to that literature. Being a reviewer for the journal is, in many ways, a proxy for expertise. When there are twice as many male "experts" as female "experts" appearing in the *AHR* it imbues our understanding of what constitutes historical expertise.

The second finding we arrived at when examining the gender of *AHR* reviewers was a better understanding of who reviews which book authors. Once again, the methodology reveals patterns that we might suspect but nonetheless tends to stay hidden from direct sight. About three times as many men as women write reviews of male-authored books in the *AHR*. In the case of female-authored books, the ratio of female-to-male reviewers is closer to 50/50. In short, women are far more likely to write reviews of other women, while men tend to gravitate more towards male authors---who are, of course, already over-represented in the *AHR*.

@@

The findings related to *AHR* reviewers help illustrate the ways in which gender inequity in the profession is so insidious. It is not always an overt case of deciding to award a prestigious prize to a male historian; it can be as seemingly innocuous as deciding to write a book review of a male author's book. Countless individual actions like these constitute the core relationships that knit together the profession, and too often these tiny actions benefit men more than women: chairing a panel of mostly male historians, taking on a new male advisee, or assigning mostly male-authored books on a syllabus. On their own they might appear benign---one might argue that male historians simply happen to be interested in similar research topics---but in aggregate they contribute to a broader professional landscape that puts female historians at a disadvantage.

One 2013 analysis of 2,500 recent history PhDs found that "gender played little role in employment patterns across particular professions and industries" [Wood and Townsend 2013]. Our own analysis reveals a different story, at least in terms of getting a book reviewed in the field's most prestigious journal. The *American Historical Review* continues to publish twice as many reviews of male-authored books as female-authored books, and those same reviews are more than twice as likely to be written by a man than a woman. This disparity is put into even sharper relief when set against the much smaller gender gap between female and male historians who complete dissertations. Women might be getting hired at comparable rates to men, but discriminatory gatekeeping remains alive and well in the profession. The gender package is one more tool to uncover those practices.

@@

The digital turn has made large datasets widely available, yet in the midst of abundance many humanists continue to deal with the problem of incompleteness and scarcity within their sources. Our method uses *abundant* existing datasets to infer additional information from *scarce* datasets. Its major contribution is to incorporate a historical method, one that takes into account how gendered naming practices have changed over time. This temporal approach provides much improved results over simpler, anachronistic, or ahistorical lookup methods. The gender package is, of course, far from perfect. By relying on historical datasets collected by state agencies, it reinforces a gender binary and larger power structure that obscures how people may identify themselves and how they experience the world. To borrow a phrase from Miriam Posner, the Gender package does not necessarily meet the "radical, unrealized potential of digital humanities" [Posner 2015]. But it might.

The gender package is entirely open-source, and we have provided detailed documentation for its code and underlying data so that researchers who wish to build upon it can modify the package accordingly. To take one imagined scenario: Facebook allows its users to select from a spectrum of gender identities or to write in their own prefered identity. Because these identities are associated with the user's first name and date of birth, this is exactly the kind of historical dataset that could be incorporated into the gender package. Doing so would replace the package's simplistic categories of "Male" and "Female" with more detailed categories like "Female to Male", "Bigender", or "Cisgender Female". Although this imagined scenario would create its own set of challenges, it would represent a major improvement in how the Gender package dealt with the complexity and fluidity of gender identity.^[This imagined scenario runs into its own set of problems. Facebook's business model makes it unlikely to grant researchers open access to its proprietary and closely-guarded datasets. And even if it did grant access to this data, Facebook collected it for particular purposes and in particular ways. Just as government agencies like the Social Security Administration collect data to serve the purposes of the American state, Facebook's data collection more accurately reflects the priorities of a global corporation rather than the lived experience of its users.] By building as much flexibility and transparency into the method as possible, we hope that others can continue to make just these kinds of improvements.
